Cybersecurity Impacts of the Covid-19 Pandemic in Italy

### 2021-03-31 article.tex
Small edits

The OAD 2020 survey
\footnote{The report is written in Italian,
  only the Executive Summary is in English: 186 A4 pages, 148 images and graphics, 11 Chapters
  (147 A4 pages) and 9 Attachments (39 A4 pages).}~\cite{oad20}
took place during the
Covid-19 pandemic and covered the entire 2019 and the first quarter
of 2020, when the pandemic exploded in Italy. The comparison of data from 2019 to that
from the first quarter of 2020, provided by the same set of respondents,
has statistical relevance and provides a 
clear indication of the Covid-19 pandemic on the intentional cyberattacks in Italy.
This pandemic has triggered a wide range of cyberattacks, mainly 
caused by the sudden, and partly unexpected, passage to working from home
and to a strong use of every type of IT service on the Internet, as a consequence
of the mobility lockdowns imposed by the Italian authorities. The resulting respondents pool
covers all the production sectors (Fig.~\ref{fig:9}), including public administration,
even if the majority of the respondents companies belong to the ICT sector (30.8\%).
However, many of them are represented by low percentages. For this reason, not having sufficiently
representative samples, we have not considered particular data analysis and correlations by sectors.
The 2020 pool is well balanced for the size of the organizations, in terms
of employees, between those below 250 and the largest ones. It must be taken into account,
as reported by ISTAT~\cite{istat21}, that 99.91\%
of enterprises in Italy are SMEs, with up to 250 employees, and 95\% have fewer than ten employees.
The OAD 2020 survey is therefore able to consider small and 
tiny organizations, that are the vast majority in Italy, and that usually
are not considered in other cybersecurity surveys: 57.5\% of the OAD 2020 respondents
belong to structures with less than 250 employees and, of these, 22.1\% have fewer than ten
employees. The distribution of annual respondents' turnover also reflects the turnover of SMEs, as shown in Fig.~\ref{fig:7}. Regarding geographical distribution (Fig.~\ref{fig:8}), the responders' companies and organizations are located for the most part only on Italian territory (77.9\%), predominantly in northern Italy (43.3\%).

@@

\begin{figure}
	\centering
	\includegraphics[width=0.90\textwidth]{pictures/fig7.png}
	\caption{Last annual available turnover of responders' companies and organizations~\cite{oad20}. 
		 It should be noted that almost 20\% of the respondents cannot or do not want provide the turnover class to which they fall, stating that the confidentiality policies prevent to indicating a turnover class, even in an investigation completely anonymous.
	}
	\label{fig:7}
\end{figure}

@@

Regarding the geographical distribution of attacks, Table.~\ref{tab:table4} shows
that occasional and infrequent attacks are evenly distributed throughout the territory in 2019,
with systematic attacks focused on northern Italy (where the companies with the highest turnover are present). 
While, Table.~\ref{tab:table5} shows that already in the first months of 2020 there were several attacks.

@@

	\caption{Distribution of digital attacks (detected by the respondents) wrt. to geographic area in 2019. The percentages of central Italy are 
	very different from the others because the organizations responding to that geographical area were numerically lower than the rest 
	of the survey participant (Fig.\ref{fig:8}).}									
	\label{tab:table4}									
\end{table}

@@

	\caption{Distribution of digital attacks (detected by the respondents) wrt. to geographic area in 1Q-2020.}									
	\label{tab:table5}									
\end{table}

The impact of the pandemic on the cybersecurity in Italy is also highlighted by the data
provided by the Italian Postal and Telecommunication Police, that shows the situation for the
critical infrastructures (Table~\ref{tab:table1}), for the financial cyber-crimes
(Table~\ref{tab:table2}) and for the cyber-terrorism (Table~\ref{tab:table3}). 
Fraudulent transactions all Italyin 2019 were blocked for a total value of \euro 21.3 million and \euro 18 million were recovered; in the first quarter of 
2020, fraudulent transactions for \euro 20.2 million were blocked,
practically reaching in four months the amount
blocked in the full twelve months of 2019, and \euro 8 million have been
recovered. This is an indication of the increase in attacks on financial transactions
and services due to the very large use of these online services, caused by the mobility reduction
measures consequent of the Covid-19 pandemic.

>>>>

The OAD 2020 survey
\footnote{The report is written in Italian,
  only the Executive Summary is in English: 186 A4 pages, 148 images and graphics, 11 Chapters
  (147 A4 pages) and 9 Attachments (39 A4 pages).}~\cite{oad20}
took place during the
Covid-19 pandemic and covered the entire 2019 and the first quarter
of 2020, when the pandemic exploded in Italy. The comparison of data from 2019 to that
from the first quarter of 2020, provided by the same set of respondents,
has statistical relevance and provides a 
clear indication of the Covid-19 pandemic on the intentional cyberattacks in Italy.
This pandemic has triggered a wide range of cyberattacks, mainly 
caused by the sudden, and partly unexpected, passage to working from home
and to a strong use of every type of IT service on the Internet, as a consequence
of the mobility lockdowns imposed by the Italian authorities. The resulting respondents pool
covers all the production sectors (Fig.~\ref{fig:9}), including public administration,
even if the majority of the respondents companies belong to the ICT sector (30.8\%).
However, many of them are represented by low percentages. For this reason, missing sufficiently
representative samples, we have not performed a particular data analysis and correlations by sectors.
The 2020 pool is well balanced for the size of the organizations, in terms
of employees, between those below 250 and the largest ones. It must be taken into account,
as reported by ISTAT~\cite{istat21}, that 99.91\%
of enterprises in Italy are SMEs, with up to 250 employees, and 95\% have fewer than ten employees.
The OAD 2020 survey is therefore able to consider small and 
tiny organizations, that are the vast majority in Italy, and that usually
are not considered in other cybersecurity surveys: 57.5\% of the OAD 2020 respondents
belong to structures with less than 250 employees and, of these, 22.1\% have fewer than ten
employees. The distribution of annual respondents' turnover also reflects the turnover of SMEs,
as shown in Fig.~\ref{fig:7}. Regarding geographical distribution (Fig.~\ref{fig:8}),
the responders' companies and organizations are mostly located in the Italian territory (77.9\%),
predominantly in northern Italy (43.3\%).

@@

\begin{figure}
	\centering
	\includegraphics[width=0.90\textwidth]{pictures/fig7.png}
	\caption{Last annual available turnover of responders' companies and organizations~\cite{oad20}. 
	  It should be noted that almost 20\% of the respondents cannot or do not want
          to provide the turnover class to which they fall, asserting that the confidentiality
          policies prevent them from indicating a turnover class,
          even in an completely anonymous investigation.
	}
	\label{fig:7}
\end{figure}

@@

Regarding the geographical distribution of attacks, Table.~\ref{tab:table4} shows
that occasional attacks were evenly distributed throughout the territory in 2019,
with systematic attacks focused on northern Italy
(where the companies with the highest turnover are present). 
While Table.~\ref{tab:table5} shows that, already in the first months of 2020,
there were several attacks.

@@

	\caption{Distribution of digital attacks (detected by the respondents) wrt.\ geographical area, in 2019. The percentages of central Italy are 
	very different from the others because the organizations responding from that geographical area were much fewer than the rest 
	of the survey participant (Fig.\ref{fig:8}).}									
	\label{tab:table4}									
\end{table}

@@

	\caption{Distribution of digital attacks (detected by the respondents) wrt.\ geographical area, in 1Q-2020.}									
	\label{tab:table5}									
\end{table}

The impact of the pandemic on the cybersecurity in Italy is also highlighted by the data
provided by the Italian Postal and Telecommunication Police, that shows the situation for the
critical infrastructures (Table~\ref{tab:table1}), for the financial cyber-crimes
(Table~\ref{tab:table2}) and for the cyber-terrorism (Table~\ref{tab:table3}). 
Fraudulent transactions were blocked in all Italy in 2019 for a total value
of \euro 21.3 million and \euro 18 million were recovered; in the first quarter of 
2020, fraudulent transactions for \euro 20.2 million were blocked,
practically reaching in four months the amount
blocked in the full twelve months of 2019, and \euro 8 million have been
recovered. This is an indication of the increase in attacks on financial transactions
and services due to the very large use of these online services, caused by the mobility reduction
measures consequent of the Covid-19 pandemic.
